---
title: "simulation-confounding"
author: "Joakim Wallmark"
date: "11/15/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simulation study

Investigate bias from missing confounders when estimating NDE and NIE.

## Simulation scenarios
variable | type | true model
---------|----- | ----------------
X(additional covariate) | continous | $X\sim gamma(8, 4.5)$
Z(exposure) | binary | $Z=I(Z*>0)$ where $Z*\sim U_0+U_1X+N(0, 1)$
M(mediator) | continous | $M\sim B_0+B_1Z+B_2X+\epsilon, \epsilon \sim N(0, 1)$
Y(outcome) | continous | $Y\sim \theta_0+\theta_1Z+\theta_2M+\theta_3ZM+\theta_4X+\omega, \omega \sim N(0, 1)$
$corr(\omega, \epsilon)=$varied (-0.5,0.5)

Table: Variables used for simulations:

* $U_0=-0.4$
* $U_1=0.01$
* $B_0=3$
* $B_1=2$
* $B_2=0.05$
* $\theta_0=5$
* $\theta_1=1$
* $\theta_2=0.5$
* $\theta_3=0.5$
* $\theta_4=0.05$

Estimated mediator model was set to the correct one. Bias was induced by changing error term correlation (between $\omega$ and $\epsilon$).

## Results

```{r echo=F}
library(ggplot2)
load("to-plot-confounding.RData")
knitr::kable(to.plot, col.names = colnames(to.plot))

ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=corr, y = true.nde, col = "NDE")) +
  geom_line(data = data.frame(to.plot), aes(x=corr, y = true.nie, col = "NIE")) +
  xlab('correlation.coef') +
  ylab('effect') +
  expand_limits(y=0)
ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=corr, y = est.nde, col = "est NDE")) +
  geom_line(data = data.frame(to.plot), aes(x=corr, y = true.nde, col = "true NDE")) +
  geom_ribbon(data = data.frame(to.plot), aes(x=corr, ymin=est.nde-nde.emp.SE, ymax=est.nde+nde.emp.SE), linetype=2, alpha=0.1) +
  xlab('correlation.coef') +
  ylab('effect')
ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=corr, y = est.nie, col = "est NIE")) +
  geom_line(data = data.frame(to.plot), aes(x=corr, y = true.nie, col = "true NIE")) +
  geom_ribbon(data = data.frame(to.plot), aes(x=corr, ymin=est.nie-nie.emp.SE, ymax=est.nie+nie.emp.SE), linetype=2, alpha=0.1) +
  xlab('correlation.coef') +
  ylab('effect')
ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=corr, y = nde.coverage, col = "NDE")) +
  geom_line(data = data.frame(to.plot), aes(x=corr, y = nie.coverage, col = "NIE")) +
  xlab('correlation.coef') +
  ylab('coverage') +
  ggtitle("Coverage of nominal 95% CI. Nominal = based on delta-SE?")
ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=corr, y = nde.emp.SE, col = "emp. SE")) +
  geom_line(data = data.frame(to.plot), aes(x=corr, y = nde.model.SE, col = "model. SE")) +
  xlab('correlation.coef') +
  ylab('SE') +
  ggtitle("Comparison of NDE SE's")
ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=corr, y = nie.emp.SE, col = "emp. SE")) +
  geom_line(data = data.frame(to.plot), aes(x=corr, y = nie.model.SE, col = "model. SE")) +
  xlab('correlation.coef') +
  ylab('SE') +
  ggtitle("Comparison of NIE SE's")
```




