---
title: "simulation summary binary"
author: "Joakim Wallmark"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simulation study

## Simulation scenarios
Simulations of bias were done with model misspecification and confounding in separate comparable scenarios. 10000 iterations (1/10th of the linear case due to computation time) with a sample size of 2000 were used for each scenario.

variable | type | true model
---------|----- | ----------------
X(additional covariate) | continuous | $X\sim gamma(8, 4.5)$
Z(exposure) | binary | $Z=I(Z*>0)$ where $Z*\sim U_0+U_1X+N(0, 1)$
M(mediator) | binary | $M=I(M*>0)$ where $M*\sim \beta_0+\beta_1Z+\beta_2X+N(0, 1)$
Y(outcome) | binary | $Y=I(Y*>0)$ where $Y*\sim \theta_0+\theta_1Z+\theta_2M+\theta_3ZM+\theta_4X+N(0, 1)$

Table: Variables used for simulations

```{r echo=F, include = F}
  library(ggplot2)
  library(ggdag)
  library(gridExtra)
```

```{r echo=F, fig.width=4.2, fig.height=4}
  dag  <- dagify(Z~X,
                 M~Z+X,
                 Y~Z+M+X)
  ggdag(dag) + ggtitle("DAG showing data generating process") + 
    theme(plot.title = element_text(size = 9, face = "bold"), 
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank())
```

### Simulation scenarios (model misspecification)
To get realistic parameter values, models were estimated using data from riksstroke from n=60353 patients. Parameter values used for simulations:

covariate.parameters = list(c(104, 8, 4.5)),
true.exposure.coefs = c(I = -3.416096, X = 0.036231),
true.mediator.coefs = c(I = 1, Z = -0.4, X = -0.008, ZX = 0),
true.outcome.coefs = c(I = 4, Z = -0.4, M = -2, ZM = corr.coef[i], X = -0.03, ZX = 0, MX = 0, ZMX = 0),

Estimated mediator model was set to the correct one. Estimated outcome model was misspecified without $ZM$ interaction: $Y*\sim Z+M+X$.

\newpage

## Results
```{r echo=F}
library(gridExtra)
load("../Data/Data-probit/to-plot-probit.reverse.i10000.n2000.RData")

int.nde = ggplot() + 
  ggtitle("True vs estimated NDE(missing int.coef.)") +
  theme(plot.title = element_text(size = 9, face = "bold"),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.position = c(0.86, 0.22)) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nde.est, col = "est NDE")) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nde.true, col = "true NDE")) +
  geom_ribbon(data = data.frame(to.plot), aes(x=interaction.coefficient, ymin=nde.est-nde.emp.SE, ymax=nde.est+nde.emp.SE), linetype=2, alpha=0.1) +
  xlab('interaction.coef') +
  ylab('effect')
int.nie = ggplot() + 
  ggtitle("True vs estimated NIE(missing int.coef.)") +
  theme(plot.title = element_text(size = 9, face = "bold"),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.position = c(0.86, 0.22)) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nie.est, col = "est NIE")) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nie.true, col = "true NIE")) +
  geom_ribbon(data = data.frame(to.plot), aes(x=interaction.coefficient, ymin=nie.est-nie.emp.SE, ymax=nie.est+nie.emp.SE), linetype=2, alpha=0.1) +
  xlab('interaction.coef') +
  ylab('effect')
grid.arrange(int.nde, int.nie, nrow =1)
```

\newpage
Total effect is biased. This differs from the situation where we used linear models:

```{r echo=F, fig.height=3}
int.total = ggplot() +
  ggtitle("Total effect(TE) (misspecified model)") +
  theme(plot.title = element_text(size = 9, face = "bold"), legend.position = "bottom") +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nde.true+nie.true, col = "True TE")) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nde.est+nie.est, col = "Estimated TE")) +
  xlab('interaction.coef') +
  ylab('total effect')
grid.arrange(int.total, nrow = 1)
```

```{r echo=F}
int.SE.NDE = ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nde.emp.SE, col = "emp. SE")) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nde.model.SE, col = "model. SE")) +
  xlab('interaction.coef') +
  ylab('SE') +
  ggtitle("Comparison of NDE SE's")+
  theme(plot.title = element_text(size = 9, face = "bold"),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 7))
int.SE.NIE = ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nie.emp.SE, col = "emp. SE")) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nie.model.SE, col = "model. SE")) +
  xlab('interaction.coef') +
  ylab('SE') +
  ggtitle("Comparison of NIE SE's")+
  theme(plot.title = element_text(size = 9, face = "bold"),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 7))
grid.arrange(int.SE.NDE, int.SE.NIE, nrow = 1)
```

\newpage

```{r echo=F, fig.width=5, fig.height=3}
cov.int = ggplot() + 
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nde.coverage, col = "NDE")) +
  geom_line(data = data.frame(to.plot), aes(x=interaction.coefficient, y = nie.coverage, col = "NIE")) +
  xlab('interaction.coef') +
  ylab('coverage') +
  ggtitle("Coverage of nominal 95% CI") +
  theme(plot.title = element_text(size = 9, face = "bold"), legend.position = "bottom")
grid.arrange(cov.int, nrow = 1)
```
